Fast Sorting on Modern Computers

### 2005-06-09 tech_report/docs/method.tex
inputs are now the single tiled mergesort ones moved paper to tech_report to allow new paper directory to actually be for the paper

\label{method}
This section discusses the method and framework used to complete this research.
The first section discusses the steps involved in creating a testing framework,
programming the sorts and getting results. The second section discusses tools
that I used or created to do this.

@@

Early in the project, it was decided that exceeding the bounds of an array
during a sort was to be forbidden. No serious sorting library or application
could acceptably work in this manner, and any results obtained while allowing
this were only of academic interest. It isn't possible, from a normal run of a
program, to determine if an arrays bounds are exceeded slightly, though exceeding
the bounds by a large amount will result is a segmentation fault. Because of
this, Valgrind (see Section \ref{Valgrind}), was used to report these errors.

@@

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc 2.63 - with related tools such as a linker and
assembler.  Several simulators are then provided: \cc{sim-fast} only counts
instruction, quickly; \cc{sim-cache} simulates a cache and counts level 1 and
level 2 cache misses; \cc{sim-bpred} simulates a branch predictor and counts
branch hits and misses. Other simulators are provided, but they were not used in
this project.

@@

The main program provided the framework for the project. Imaginatively called
\cc{main.c}, this program provided the means for testing and (roughly)
timing each sort. Figure \vref{Key sections of Main.c} contains this code.

@@

Perfex was an alternative to using SimpleScalar. By using its hardware
registers, actual real world results could have been measured. Using perfex
instead of SimpleScalar was an option reviewed at the start, but the difficulty
involved in setting up a machine to use it was considerable. Additionally,
because perfex measures results on a real machine, only one type of cache and
branch predictor could have been measured. Since access was
provided late in the project, limited measurements were taken to supplement the
SimpleScalar results.

@@

Algorithm N is described in \cite{Knuth98}. A flow control diagram and step by
step guide is provided, but very little detail of how the sort actually works is
included. Therefore, the first step was to understand how this sort worked. This
was done by first analysing the flow control of the algorithm. The steps of the
sort are described in assembly, and while they can be written in C, the result
is a very low-level version. I attempted to rewrite the algorithm without
\cc{goto} statements and labels, but instead using standard flow control
statements such as \cc{if}, \cc{while} and \cc{do-while}. Despite the time spent
on this, it was impossible to faithfully reproduce the behaviour without
duplicating code.

@@

As well as the number of cycles, Pentium 4 performance registers can also provide
most of the same information as SimpleScalar, with different parameters. For
example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
cache lines were used in SimpleScalar.  The addition of real world data would
strengthen the results from this project.

>>>>

\label{method}
This section discusses the method and framework used to complete this research.
The first section discusses the steps involved in creating a testing framework,
programming the sorts and getting results. The second section discusses tools
that were used or created to do this.

@@

Early on, it was decided that exceeding the bounds of an array
during a sort was to be forbidden. No serious sorting library or application
could acceptably work in this manner, and any results obtained while allowing
this were only of academic interest. It isn't possible, from a normal run of a
program, to determine if an arrays bounds are exceeded slightly, though exceeding
the bounds by a large amount will result is a segmentation fault. Because of
this, Valgrind (see Section \ref{Valgrind}), was used to report these errors.

@@

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc 2.63 - with related tools such as a linker and
assembler.  Several simulators are then provided: \cc{sim-fast} only counts
instruction, quickly; \cc{sim-cache} simulates a cache and counts level 1 and
level 2 cache misses; \cc{sim-bpred} simulates a branch predictor and counts
branch hits and misses. Other simulators are provided, but they were not used in
our tests.

@@

The main program provided the framework for the development of the sorts.
Imaginatively called \cc{main.c}, this program provided the means for testing
and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains
this code.

@@

Perfex was an alternative to using SimpleScalar. By using its hardware
registers, actual real world results could have been measured. Using perfex
instead of SimpleScalar was an option reviewed at the start, but the difficulty
involved in setting up a machine to use it was considerable. Additionally,
because perfex measures results on a real machine, only one type of cache and
branch predictor could have been measured. We have provided limited
measurements to supplement the SimpleScalar results.

@@

Algorithm N is described in \cite{Knuth98}. A flow control diagram and step by
step guide is provided, but very little detail of how the sort actually works is
included. Therefore, the first step was to understand how this sort worked. This
was done by first analysing the flow control of the algorithm. The steps of the
sort are described in assembly, and while they can be written in C, the result
is a very low-level version. An attempt was made to rewrite the algorithm
without \cc{goto} statements and labels, instead using standard flow control
statements such as \cc{if}, \cc{while} and \cc{do-while}. Despite the time spent
on this, it was impossible to faithfully reproduce the behaviour without
duplicating code.

@@

As well as the number of cycles, Pentium 4 performance registers can also provide
most of the same information as SimpleScalar, with different parameters. For
example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
cache lines were used in SimpleScalar. The addition of real world data would
strengthen our results.
